Characterizing subarctic European vegetation for land surface modeling of ozone pollution and climate risk

### 2021-10-03 VegOzone.tex
Restructure and elaborate on conclusions

\textbf{TODO: Decide on terms acclimation vs parameterization etc} We have studied the effect of model acclimations of some common vegetation types to a subarctic climate on phytotoxic ozone uptake. The comparison between 2018 and 2019 conditions allowed us to consider these acclimations and their influence on vegetation sensitivity to \chem{O_3} in light of future changes of key environmental variables as may occur under climate change (e.g. increase of frequency and extent of heatwaves). We found that conditions for ozone formation were more favorable in the 2018 growing season than in 2019, with 2018 being significantly warmer and less cloudy in spring and early summer. Accordingly, peak ozone concentrations occurred more frequently and at higher levels in 2018. This was particularly a result of the extended heatwave in spring and early summer and associated, extensive wildfires in Central Sweden (in particular in mid--end July). \textbf{TODO: Check explanation regarding significance.}

We found that the length and timing of the growing season are crucial for risk assessment based on accumulated ozone flux metrics ($\mathrm{POD_y}$). Based on a MODIS photosynthesis product, we estimated the growing season for coniferous trees in 2018 to be at least one month shorter than in 2019. The found start of growing season in 2018 ($122\,\unit{doy}$) and 2019 ($106\,\unit{doy}$) are within given ranges by empirical observations \citep{TB:Kolari2007,IVL:Karlsson2018}. Based on gridded observational temperature and snow depth, we found a half a month shorter growing season for deciduous trees and perennial grassland in 2018 compared to 2019, respectively. With respect to ongoing climate change, a clear positive trend in length ($5.2\,\unit{d\,decade^{-1}}$) almost equally distributed between earlier start ($2.9\,\unit{days\,decade^{-1}}$) and later end ($2.3\,\unit{d\,decade^{-1}}$) could be deducted from the thermal growing season at Svanhovd (Appendix Fig.~\ref{fig:greening_season_change_Svanvik}). \textbf{TODO: Move S6 into appendix.}

Ozone peak concentrations were observed more frequently in 2018 than in 2019. Nevertheless, \chem{POD_1} values were very similar for both years. This suggests that the overall length of the growing season is more important in risk assessment using flux-based metrics than episodes of peak concentrations. These peak \chem{[O_3]} often coincide with environmental conditions that limit ozone uptake the model. However, acute cellular damage caused by very high episodic ozone fluxes are not accounted for in $\mathrm{POD_y}$. That might explain why ozone damage was observed at Svanhovd to a larger extent in 2018 than in 2019. As pointed out by \citet{AE:Musselman2006}, a susceptibility to ozone depending on leaf age in the associated damage functions might give a different picture. Forbs and perennial grassland have been found to be more susceptible in the reproductive state \citep{EP:Bassin2004}.

To estimate potential biases in the assessment of ozone-induced risk on Fennoscandic vegetation, we defined hypothetical temperature response functions representing different levels of assumed acclimation to cold climates. We assume northern Fennoscandia vegetation to be more tolerant towards cold temperatures and less tolerant to heat. Based on probability density functions (PDF) of observed light and temperature, we assume that an acclimation ought to increase the enclosed area between the respective response functions and the PDF. These hypothetical parameterisations of species or PFTs still need verification by experimental data.

In summary, our key findings are:
\begin{enumerate}
   \itemsep0pt
 \item the bespoke parameterizations for subarctic species are important (especially concerning light and temperature and the climate effect on the growing season) to be able to assess \chem{O_3} uptake,
 \item the parameterizations defined in the mapping manual for European regional risk assessment appear to not adequately capture the physiological responses of subarctic physiology that are important determinants of \chem{[O_3]} sensitivity and are likely to substantially underestimate risk, and
 \item that the climatic conditions promoting ozone damage (particularly extended heatwaves and the timing of the growing season concerning the ozone spring peak) and which are likely to become more frequent under climate change are likely to further enhance the sensitivity of subarctic vegetation to ozone.
\end{enumerate}

However, the decline of the ozone spring peak is partly caused by dry deposition of ozone on vegetation. It, therefore, remains unclear whether an earlier start of the growing season has the potential to reduce the ozone spring peak concentration compared to present-day and hence reduce the damage risk of ozone in the future. 
We have not assessed the effect of the plants' response to VPD here, but our climatological data show that conditions at Svanhovd are shifting towards a more VPD limited regime.

Based on the methods described in \citet{ICP:MappingManual2017}, we estimated biomass reductions due to ozone uptake for both 2018 and 2019 and found significant reductions for deciduous trees ($15.5-17.4\,\unit{\%}$) and perennial grassland (above ground, $12.4-13.7\,\unit{\%}$) with respect to estimates based on $\mathrm{POD_1}$ from default parameterizations. For coniferous trees, the adjusted parameterizations are comparable with the default parameterization estimates and are contained within the range of systematic uncertainties deduced from our bespoke response function ($2.5-3\,\unit{\%}$). These results have to be treated with care because the biomass reduction functions were not established based on subarctic species.

At the same time, intra-species variability even at the same location is typically non-negligible \citep{EP:Bassin2004,Amb:Girgzdiene2009}. This may explain contradictory results regarding the ozone sensitivity of natural vegetation in northern Europe. While \citet{FS:Subramanian2014} reported a modeled reduction of net primary production in coniferous ($4.3-15.5\,\unit{\%}$) and birch ($1.4-4.3\,\unit{\%}$) under elevated ozone in Sweden, Lithuanian forest deciduous trees were observed to be more susceptible to visible damage by ozone than Norway spruce \citep{Amb:Girgzdiene2009}. Open-top chamber (OTC) experiments performed in northern Finland where Scots pine and downy birch were exposed to elevated ozone concentrations showed biomass reductions of both downy birch and mountain birch seedlings after one season of exposure \citep{Amb:Manninen2009}. The leaves of the mountain birch in the high ozone treatments showed necrotic stipples as early as June. As mountain birch very rapidly develops during spring and early summer, compensation for damaged leaves is difficult for this species. In contrast to mountain birches, northern Scots pine of local origin only showed invisible needle-level \chem{O_3} impacts after three years of exposure. There is no field data comparing the ozone sensitivity of birches and Scots pine from northern Europe.
In our study, we do not explicitly account for this natural distribution of individual plant traits, but the chosen hypothetical, bespoke response functions might reflect this diversity within a given species or PFT. The resilience of northern Fennoscandia plant communities to climate change and predictions in this regard strongly depends on both the existing range of acclimations as well as the actual type of acclimation.

Cumulative ozone uptake has been reported to reduce stomatal conductance significantly and independently from photosynthesis \citep[e.g.]{BGS:Lombardozzi2012}, with photosynthesis being more strongly affected. This might be due to competing effects, as an ozone-induced sluggishness \citep{SR:Hoshika2015} renders stomata unable to fully close and leads to higher stomatal conductance than expected and thus higher ozone uptake as well as transpiration. At present, the MM version of the $\mathrm{DO_3SE}$ model does not account for any ozone-inflicted reduction of stomatal conductance throughout the growing season and is not coupled to photosynthesis. Hence, the presented $\mathrm{POD_y}$ may be regarded as an estimated upper limit.\\

Beyond the risk assessment of ozone-induced damage, our hypothetical bespoke temperature response functions, once verified by observations, can have important implications on land-surface modeling in global models, where problems in productivity of species especially in the Arctic regions occur due to PFTs which are not suitable for the climate. We pointed out, with respect to default perennial grassland parameterization, that these perennial grasslands did not show any substantial stomatal conductance in 2019 characterized as a normal year. In terms of GPP this would mean unrealistically low carbon sequestration. Automation of the here proposed PDF-based acclimation using machine learning techniques could overcome these issues in the future.

>>>>

We have studied the effect of bespoke parameterizations of common vegetation types (coniferous and deciduous trees, perennial grassland) to a subarctic climate on $\mathrm{POD_y}$. The comparison between meteorological conditions in 2018 and 2019 and their divergence from climatology allowed us to assess the influence of key environmental variables such as temperature, PPFD, and precipitation on vegetation susceptibility to \chem{O_3} damage in light of future changes as may occur under climate change. We found that conditions for ozone formation were more favorable in the 2018 growing season than in 2019, with 2018 being significantly warmer and less cloudy in spring and early summer. Accordingly, peak ozone concentrations occurred more frequently and at higher levels in 2018. This was particularly a result of the extended heatwave in spring and early summer and associated, extensive wildfires in Central Sweden. However, ozone concentrations during the growing season were not significantly enhanced and \chem{POD_1} values were very similar for both years.

Based on a MODIS photosynthesis product \citep{MODIS_PSN}, we estimated the growing season for coniferous trees in 2018 to be at least one month shorter than in 2019. The deduced start of growing season in 2018 ($122\,\unit{doy}$) and 2019 ($106\,\unit{doy}$) are within given ranges by empirical observations \citep{TB:Kolari2007,IVL:Karlsson2018}. We determined the start of the growing season for deciduous trees based on gridded observational temperature and for perennial grassland on snow depth data. For perennial grassland, we assumed a one month latency period after snow melt which is supported by observational data from Finland \citep{FCR:Korhonen2018}. With respect to ongoing climate change, a clear positive trend emerged in length ($5.2\,\unit{d\,decade^{-1}}$) of the growing season which is almost equally distributed between earlier start ($2.9\,\unit{days\,decade^{-1}}$) and later end ($2.3\,\unit{d\,decade^{-1}}$) (Appendix Fig.~\ref{fig:greening_season_change_Svanvik}).

This suggests that the overall length of the growing season is more crucial in risk assessment when using flux-based metrics than episodes of peak concentrations as these peak \chem{[O_3]} often coincide with environmental conditions that limit ozone uptake in the $\mathrm{DO_3SE}$ model. Visible damage caused by peak ozone uptake and a dependence of ozone sensibility on phenology are not accounted for in $\mathrm{POD_y}$. But open-top chamber (OTC) experiments performed in northern Finland where Scots pine and downy birch seedlings were exposed to elevated ozone concentrations showed a reduction in biomass and reproduction with visible damage explicitly attributed to peak \chem{O_3} concentrations and fast phenological development at high growth rate \citep{Amb:Manninen2009}. In particular, forbs and perennial grassland have been found to be more susceptible to ozone-induced damage in the reproductive state \citep{EP:Bassin2004}. That might explain why we observed ozone damage to a larger extent in 2018 than in 2019. An associated damage function to ozone depending on leaf age as proposed by \citet{AE:Musselman2006} or phenological state might improve the predictions.

At the same time, intra-species variability even at the same location is typically non-negligible \citep{EP:Bassin2004,Amb:Girgzdiene2009} but not accounted for in the models. This may also explain some contradictory results regarding ozone sensitivity of natural vegetation in northern Europe. \citet{FS:Subramanian2014} reported a higher, modeled reduction of net primary production in conifers ($4.3-15.5\,\unit{\%}$) than birch ($1.4-4.3\,\unit{\%}$) under elevated ozone in Sweden, while \citet{Amb:Girgzdiene2009} observed more visible damage on deciduous trees than on Norway spruce in Lithuania. But visible damage and biomass reduction may not occur at the same rate and hence may not be interchangeable. In our study, we do not explicitly account for this natural distribution of individual plant traits, but our bespoke response functions reflect the diversity within a given species. In particular, we assumed two tolerance levels to cold temperatures leading to a higher efficiency in photosynthesis at cool temperatures than for central European species. To this end, we computed a PDF from observed temperature at the NIBIO environment centre Svanhovd (May--September, 1992--2000). We adjusted the parameters in the temperature response function of stomatal conductance to increase the enclosed area with the PDF. Assuming a proportionality between $g_\mathrm{sto}$ and $A_\mathrm{n}$, this is equivalent to optimizing carbon uptake. Similarly, we deduced a PPFD PDF and varied the light response function to test a higher or lower extent of stomatal opening at low light intensities. These parameterizations are hypothetical and need verification by experimental data. We found that soil water potential under 2018/19 meteorological conditions was negligible when considering the bespoke growing season. We have not assessed the effect of the plants' response to VPD in detail, but our climatological data indicated that conditions at Svanhovd are shifting towards a more VPD limited regime. Our key findings are: (1) Our bespoke parameterizations for subarctic species are better suited to describe vegetation adapted to subarctic conditions for instance allowing for stomata opening during a larger part of the growing season, (2) the parameterizations defined in the mapping manual for European regional risk assessment appear to not adequately capture the physiological responses of subarctic vegetation that are important determinants of \chem{[O_3]} susceptibility and are likely to underestimate ozone-damage risk, and (3) climatic conditions that are promoting a longer growing season are likely to increase the ozone-damage risk in flux-based metrics.

In this regard, vegetation might become more exposed to higher ozone concentrations in early phenological stages due to an overlap with ozone spring peak \citep{EP:Karlsson2007}. However, the decline of this ozone spring peak is partly caused by uptake of vegetation. It, therefore, remains unclear whether an earlier start of the growing season will increase the exposure of vegetation to ozone or lead to a compensation due to an earlier decline of the peak ozone in the future. \citet{ESPR:Hayes2021} pointed out that the highest susceptibility of vegetation to future ozone variability is likely to occur in summer when vegetation is most productive. The actual resilience of northern Fennoscandia plant communities to climate change and its predictions strongly depend on both the range of existing acclimations as well as the actual type.

Following the methods described in \citet{ICP:MappingManual2017}, we estimated biomass reductions due to ozone uptake for both 2018 and 2019 and found substantial reductions for deciduous trees ($15.5-17.4\,\unit{\%}$) and perennial grassland (above ground, $12.4-13.7\,\unit{\%}$) in both years. These reductions are higher than for the default parameterizations, $11.2\,\unit{\%}$ (deciduous) and $9.6\,\unit{\%}$ (perennial grassland). For coniferous trees, the estimated reduction in biomass differs insignificantly between our bespoke parameterizations ($2.5-3\,\unit{\%}$) and the default ($2.31\,\unit{\%}$). This indicates a higher modeled robustness to ozone damage in coniferous trees. All these results have to be treated with care because the biomass reduction functions were established based on species in less extreme climates. In particular, CLs for deciduous and coniferous trees were breach in all years and fora ll parameterizations. For perennial grassland, the CL was not breached if a bespoke GS was assumed, but the mapping manual defined CL of $10\,\unit{\%}$ loss in biomass might not be transferable to subarctic regions with considerable constrains on productivity leading to more severe economical consequences.

Beyond the risk assessment of ozone-induced damage, our bespoke temperature response functions, once verified by observations, can have important implications on land-surface modeling in global models, where problems in productivity of species especially in the Arctic regions occur due to PFTs which are not suitable for the climate. We pointed out, with respect to default perennial grassland parameterization, that these perennial grasslands did not show any substantial stomatal conductance in 2019 characterized as a normal year. In terms of GPP this would mean unrealistically low carbon uptake. Automation of the here proposed PDF-based acclimation using machine learning techniques could overcome these issues in the future.
